For Induction-Induction, Induction is Enough

### 2019-11-11 paper/paper.tex

We first define a \emph{theory of IIT signatures} in which every context is a
valid IIT signature (Section \ref{sec:theory_of_signatures}). We present this
type theory as an algebraic \cite{ttintt} theory, specifically as a quotient
inductive-inductive type (QIIT). In Section \ref{sec:ambroise} we show that this
QIIT can be constructed only using inductive types.

@@

\section{Constructing IITs from the Theory of IIT Signatures}
\label{sec:andras}

We construct all IITs describable by the theory of IIT signatures through the
term model construction of \cite{Kaposi:2019:CQI:3302515.3290315}. There, it is
shown that from the syntax of the theory of signatures for finitary QIITs, one
can construct any particular finitary QIIT. The idea is that for a signature
$\Gamma$, the initial algebra can be built from sets of terms of the form
$\Tm\,\Gamma\,(\El\,a)$.

In this section, we only need to check that the signatures in
\cite{Kaposi:2019:CQI:3302515.3290315} and the term model construction restricts
to IITs, i.e.\ that from a syntax for the theory of IIT signatures all finitary
IITs are constructible.

@@

The current work only concerns finitary IITs. An extension would be to also
allow infinitely branching inductive types such as W-types. This would first
require giving semantics for infinitary IITs (to our knowledge there is no
previously published semantics that we can borrow), and also giving a term model
construction analogously to finitary QIITs. These steps seem feasible. However,
it seems to be more difficult to construct the syntax of infinitary IIT
signatures without using quotients. The reason is that such syntax would not be
strictly restricted to neutral terms: we would need $\lambda$-abstraction and
$\beta\eta$-rules for infinitary $\Pi$ types, in order to allow a term model
construction for infinitary IITs. A definition for normal preterms and typing
judgments on them may still be possible, but it appears to be much more
complicated than before (the current authors attempted this without conclusive
success).

>>>>

We first define a \emph{theory of IIT signatures} in which every context is a
valid IIT signature (Section \ref{sec:theory_of_signatures}). We present this
type theory as an algebraic \cite{ttintt} theory, specifically as a quotient
inductive-inductive type (QIIT). In Section \ref{sec:ambroise} we show that this
QIIT can be constructed only using inductive types. In Section
\ref{sec:constructingiits} we describe the construction of IITs from the syntax
of IIT signatures.

@@

\section{Constructing IITs from the Theory of IIT Signatures}
\label{sec:constructingiits}

We construct all IITs describable by the theory of IIT signatures through the
term model construction of \cite{Kaposi:2019:CQI:3302515.3290315}. There, it is
shown that from the syntax of the theory of signatures for finitary QIITs, one
can construct any particular finitary QIIT. The idea is that for a signature
$\Gamma$, the initial algebra can be built from sets of terms of the form
$\Tm\,\Gamma\,(\El\,a)$. For example, taking the signature of natural numbers:
\[
  \Gamma := \cdot\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \ra \El\,N
\]
we have that $\Tm\,\Gamma\,(\El\,N)$ is the set of natural numbers up to isomorphism,
since the only way to construct these terms is by using $s$ and $z$ from $\Gamma$.

In this section, we only need to check that the signatures and term model
construction in \cite{Kaposi:2019:CQI:3302515.3290315} restrict to IITs,
i.e.\ that from a syntax for the theory of IIT signatures all finitary IITs are
constructible.

@@

The current work only concerns finitary IITs. An extension would be to also
allow infinitely branching inductive types such as W-types. This would first
require giving semantics for infinitary IITs (to our knowledge there is no
previously published semantics that we can borrow), and also giving a term model
construction analogously to finitary QIITs. These steps seem feasible. However,
it seems to be more difficult to construct the syntax of infinitary IIT
signatures without using quotients. The reason is that such syntax would not be
strictly restricted to neutral terms: we would need $\lambda$-abstraction and
$\beta\eta$-rules for infinitary $\Pi$ types, in order to allow a term model
construction for infinitary IITs. A definition for normal preterms and typing
judgments on them may still be possible, but it appears to be much more
complicated than before (the current authors have attempted this without
conclusive success).

### 2019-11-11 paper/universaliit.tex

This approach is an alternative version of Streicher's method for interpreting
preterms in an arbitrary model\cite{streichersemantics}. Streicher first defines
a family of partial maps from the presyntax to a model, then shows that the maps
are total on well-formed input. We have found that our approach is significantly
easier to formalise. Too see why, note that the right notion of partial map in
type theory, which does not presume decidable definedness, is fairly
heavyweight:
\[
  \mathsf{PartialMap}\,A\,B := A\ra (P : \mathsf{hProp})\times(P \ra B)
\]
Here $\mathsf{hProp}$ denotes a type which is propositional. In the above
definition, we notice an opportunity for converting a fibered definition of a
type family into an indexed one; if we drop the propositionality for $P$ for the
time being, we may equivalently return a family indexed over $B$, which is
exactly just a relation $A\ra B\ra\Set$. Then, in our approach, we recover
uniqueness of $P$ through the functionality requirement on the $A\ra B\ra\Set$
relation, and totality by already assuming well-formedness in the syntax. In type
theory, using indexed families instead of display maps is a common convenience,
since the former are natively supported in type theory, while the latter require
carrying around auxiliary propositional equalities.

>>>>

This approach is an alternative version of Streicher's method for interpreting
preterms in an arbitrary model\cite{streichersemantics}. Streicher first defines
a family of partial maps from the presyntax to a model, then shows that the maps
are total on well-formed input. We have found that our approach is significantly
easier to formalise. Too see why, note that the right notion of partial map in
type theory, which does not presume decidable definedness, is fairly
heavyweight:
\[
  \mathsf{PartialMap}\,A\,B := A\ra (P : \mathsf{hProp})\times(P \ra B)
\]
Here $\mathsf{hProp}$ denotes a type which is propositional. In the above
definition, we notice an opportunity for converting a fibered definition of a
type family into an indexed one; if we drop the propositionality for $P$ for the
time being, we may equivalently return a family indexed over $B$, which is
exactly just a relation $A\ra B\ra\Set$. Then, in our approach, we recover
uniqueness of $P$ through the functionality requirement on the $A\ra B\ra\Set$
relation, and totality by already assuming well-formedness of $A$. In type
theory, using indexed families instead of display maps is a common convenience,
since the former are natively supported in type theory, while the latter require
carrying around auxiliary propositional equalities.
